Slitasje og kjørespor

### 2023-01-20 DRAFT_slitasjeOgKjørespor.Rmd
doc: Add list of steps

@@

#### Nature type mapping

@@

```{r, fig.cap="Slitasje scores."}
temp <- naturetypes_wide %>%
  group_by(slitasje) %>%
  summarise(sum = n())

@@

#### ANO

@@

##### Combine Naturtype data and ANO

@@

###### Points to polygons
The ANO data is point data, and the nature type data is multipolygon.
Because later we want to save this as a shape file, we cannot have a
mixed class type. I will therefore convert the points to polygons. I use
the area column to calculate a radii that gives that area.

@@

### Scaled indicator values

@@

### Homogeneous impact areas
I want to use the [Homogeneous Impact Areas] (HIA) to define smaller regions into which I can extrapolate the indicator values.
This data is generated by discretizing the Norwegian Infrastructure Index. I refer to the ordinal values of the
four impact area classes as their *Human Impact Factor* (HIF).

@@

I also want to split these four HIA classes based on their region. To do this I need the two layers to have the same crs
```{r}
st_crs(HIA) == st_crs(regions)
```

```{r}
HIA <- st_transform(HIA, 25833)
```

@@

Since the two layers are completely overlapping, we can get the intersections
```{r, eval=F}
HIA_regions <- eaTools::ea_homogeneous_area(HIA,
                             regions,
                             keep1 = "infrastructureIndex",
                             keep2 = "region")
saveRDS(HIA_regions, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
```

@@

I now have 20 unique areas that I will, given there is data, extrapolate indicator values over: 
```{r}
HIA_regions <- HIA_regions %>%
  mutate(region_HIF = paste(region, infrastructureIndex))
length(unique(HIA_regions$region_HIF))
```

Let's calculate the areas of these polygons and compare the HIF in the five regions. 
First we need to remove some marine areas.
```{r}
HIA_regions2 <- HIA_regions %>%
  st_intersection(outline)
```

@@

### Aggregate and spread (extrapolate)
I now want to find the mean indicator value for each HIA (i.e. to aggregate) and to spread these out
spatialy to the entire HIA within each region (i.e. to extrapolate).

@@

Get map of major roads, for context
```{r}
hw_utm <- readRDS("data/cache/highways_trondheim.rds")
```

@@

Let's split the ID column to get the HIF
```{r}
slitasje_data_extrapolated <- slitasje_data_extrapolated %>%
  tidyr::separate(ID, into = c("region", "HIF"),
                  sep = " ")
slitasje_data_extrapolated$HIF <- as.numeric(slitasje_data_extrapolated$HIF)
```

@@

Split one more time, into regions:
```{r, eval=F}
slitasje_data_extrapolated2 <- ea_spread(indicator_data = slitasje_data_extrapolated,
                         indicator = meanIndicatorValue,
                         regions = regions,
                         groups = region,
                         threshold = 20)
saveRDS(slitasje_data_extrapolated2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
```

@@

But I also need to export a map of the indicator where we extrapolate the indicator values
to the homogeneous areas. This map should be interpreted with caution, since they seem
to provide explicit indicator values for areas where there is no data. This map is mainly for
aggregation purposes. I will mark this map with the prefix _homogeneous-areas_.

@@

### Sliatsje in wetlands
Prepare export
```{r}
exp_wetland <- select(slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",],
              origin,
              area,
              i_2020,
              SHAPE
              )
```

Write to file
```{r, eval=F}
st_write(exp_wetland, "indicators/indicatorMaps/slitasje/wetland_slitasje.shp",
         append = F)
```

>>>>

1.  Import [Nature type data set](#NTM) set (incl. GRUK) and [ANO data
    set](#ANO-import)

2.  Identify the [relevant](#naturtype) nature types and [subset](#NTM)
    the data

3.  Covert [ANO points to polygons](#ANO-points-to-poly)

4.  For each polygon, extract the NiN-variable that has the lowest value
    (worst-rules rule), e.g. like Fig. \@ref(fig:slitasje-naturetype).

5.  [Combine](#combine-nt-ano) data sets

6.  [Scale](#scale-slitasje-ind) the `slitasje` variable

7.  Define [homogeneous impact areas](#HIA) (HIA)

8.  [Aggregate and spread](#spread-slitasje) indicator value across HIAs
    and assessment regions

9.  Crop based on ecosystem delineation

10. Confirm relationship between infrastructure index and indicator
    values to justify the extrapolation

11. Calculate uncertainty?

12. Export indicator maps and extrapolated maps

@@

#### Nature type mapping {#NTM}

@@

```{r slitasje-naturetype, fig.cap="Slitasje scores."}
temp <- naturetypes_wide %>%
  group_by(slitasje) %>%
  summarise(sum = n())

@@

#### ANO {#ANO-import}

@@

##### Combine Naturtype data and ANO {#combine-nt-ano}

@@

###### Points to polygons {#ANO-points-to-poly}

The ANO data is point data, and the nature type data is multipolygon.
Because later we want to save this as a shape file, we cannot have a
mixed class type. I will therefore convert the points to polygons. I use
the area column to calculate a radii that gives that area.

@@

### Scaled indicator values {#scale-slitasje-ind}

@@

### Homogeneous impact areas {#HIA}

I want to use the [Homogeneous Impact Areas](#HIA) (HIA) to define
smaller regions into which I can extrapolate the indicator values. This
data is generated by discretizing the Norwegian Infrastructure Index. I
refer to the ordinal values of the four impact area classes as their
*Human Impact Factor* (HIF).

@@

Here is a view of the data zooming in on Trondheim

```{r}
myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
                ymin = 7032142.5, ymax = 7045245.27),
                crs = st_crs(HIA))
```

```{r}
HIA <- sf::st_crop(HIA, myBB)
```

Get map of major roads, for context

```{r}
hw_utm <- readRDS("data/cache/highways_trondheim.rds")
```

<!-- ERROR  marine areas still here -->
```{r}
tm_shape(HIA)+
  tm_polygons(col = "infrastructureIndex",
    title="Infrastructure index\n(modified 4-step scale)",
    palette = "viridis",
    style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")
```

I also want to split these four HIA classes based on their region. To do
this I need the two layers to have the same CRS.

@@

Since the two layers are completely overlapping, we can get the
intersections

```{r, eval=F}
HIA_regions <- eaTools::ea_homogeneous_area(HIA,
                             regions,
                             keep1 = "infrastructureIndex",
                             keep2 = "region")
saveRDS(HIA_regions, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
```

@@

I now have 20 unique areas that I will, given there is data, extrapolate
indicator values over:

```{r}
HIA_regions <- HIA_regions %>%
  mutate(region_HIF = paste(region, infrastructureIndex))
length(unique(HIA_regions$region_HIF))
```

Let's calculate the areas of these polygons and compare the HIF in the
five regions. First we need to remove some marine areas.

```{r}
HIA_regions2 <- HIA_regions %>%
  st_intersection(outline)
```

@@

### Aggregate and spread (extrapolate) {#spread-slitasje}

I now want to find the mean indicator value for each HIA (i.e. to
aggregate) and to spread these out spatialy to the entire HIA within
each region (i.e. to extrapolate).

@@

Get map of major roads, for context

```{r}
hw_utm <- readRDS("data/cache/highways_trondheim.rds")
```

@@

Let's split the ID column to get the HIF

```{r}
slitasje_data_extrapolated <- slitasje_data_extrapolated %>%
  tidyr::separate(ID, into = c("region", "HIF"),
                  sep = " ")
slitasje_data_extrapolated$HIF <- as.numeric(slitasje_data_extrapolated$HIF)
```

@@

Split one more time, into regions:

```{r, eval=F}
slitasje_data_extrapolated2 <- ea_spread(indicator_data = slitasje_data_extrapolated,
                         indicator = meanIndicatorValue,
                         regions = regions,
                         groups = region,
                         threshold = 20)
saveRDS(slitasje_data_extrapolated2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
```

@@

But I also need to export a map of the indicator where we extrapolate
the indicator values to the homogeneous areas. This map should be
interpreted with caution, since they seem to provide explicit indicator
values for areas where there is no data. This map is mainly for
aggregation purposes. I will mark this map with the prefix
*homogeneous-areas*.

@@

### Sliatsje in wetlands

Prepare export

```{r}
exp_wetland <- select(slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",],
              origin,
              area,
              i_2020,
              SHAPE
              )
```

Write to file

```{r, eval=F}
st_write(exp_wetland, "indicators/indicatorMaps/slitasje/wetland_slitasje.shp",
         append = F)
```
